Combinatorics of the geometry of Wilson loop diagrams II: Grassmann necklaces, dimensions, and denominators

### 2019-04-23 GWLDCII.tex
some progress on the GN theorem

The entries $c_{p,q}$ are non-zero real variables. The columns of $C(W)$ correspond to the vertices of $W$ listed in their linear order, and the rows correspond to the propagators of $W$. $C(W)$ is therefore only defined up to rearrangement of the rows, which will not matter in this setting since we will only interpret it as carving out a family of elements of the $|\cP| \times |V|$ real Grassmannian.

@@

From $C(W)$ we obtain the matroid $M(W)$ associated to $W$, which is simply the matroid represented by $C(W)$: its ground set is the set $V$ of vertices of $W$, and the independent sets are exactly the sets $U \subseteq V$ such that the columns of $C(W)$ indexed by $U$ are linearly independent.

It was shown in \cite{wilsonloop} that $M(W)$ is actually a positroid whenever $W$ is admissible; the matrix $C(W)$ therefore parametrizes a subspace $\Sigma(W)$ of the totally nonnegative real Grassmannian $\Gr(k, n)$.

Many geometric and combinatorial techniques exist to study the behavior of positroids in $\Gr(k,n)$; more detail on this is given in Section~\ref{sec:positroid background} below. Our perspective in this paper is to study the Wilson loop diagrams by studying the positroids they define.

@@

@@

Since Wilson loop diagrams correspond to positroids, it is natural to study the subspace of $\Gr(|\cP|, n)$ they define.

@@

\begin{dfn}\label{props inside p}
Let $W = (\cP, [n])$ be a weakly admissible Wilson loop diagram, with $p \in \cP$ supported on edges $i$ and $j$. Write $(p, <_i)$ to be the same propagator with a direction induced by the linear ordering $<_i$. Then
\begin{gather*}\cP_{in}(p,<_i) = \{ q \in \cP \ |\ (q,<_i) = (k,l) \text{ and }i \leq_i k <_i l \leq_i j \} \\
\cP_{out}(p,<_i) = \{ q \in \cP \ | \ (q,<_i) = (k,l) \text{ and } j \leq_i k <_i l  \} \;\end{gather*}
are the sets of propagators inside (resp. outside) $p$. \note{by this definition, $p \not\in \cP_{out}$}
\end{dfn}

@@

Suppose for a contradition that there exists an admissible diagram for which there exists an $i$ with $k\in I_{i\backslash\{i\}}$ and $k \not\in I_{i+1}$.  Fix $n$.  Let the triple $(W, i, k)$ be such a counterexample on $n$ vertices which is minimal with respect to the number of propagators. %, and among all those counterexamples on $n$ vertices with the same number of propagators, let $(W,i,k)$ be minimal with respect to the distance from $i$ to $k$ in the $\leq _i$ order.

@@

We have shown that $(I_1,\dots,I_n)$ is a Grassmann necklace; it remains to check that this Grassmann necklace corresponds to the positroid $M(W)$.  We need to show that:
\begin{itemize}
\item For each $i \in [n]$, $I_i$ is a basis for $M(W)$.
\item If $J$ is lexicographically smaller than $I_i$ with respect to $<_i$, then $J$ is not a basis for $M(W)$.\note{Si\^an note to self: check whether it's lex smaller or Gale smaller}
\end{itemize}
The algorithm is pairing each $j \in I_i$ with a unique propagator supported on that vertex so by Lemma~\ref{lem basis as perm} $I_i$ is a basis for $M(W)$.
%%Recall from [ref] that a $k$-subset $J \in \binom{[n]}{k}$ is a basis for $\PP_W$ if and only if it has no subset $U \subseteq J$ such that $|U| > |Prop(U)|$.  It is immediately clear from the construction that $I_i$ is a basis for each $i$: the algorithm is pairing each $j \in I_i$ with a unique propagator supported on that vertex, so $|Prop(U)| \geq U$ for all $U \subseteq I_i$.

%%If $\exists l \in J$ with $Prop(l) = \emptyset$ then $J$ is clearly not a basis, so suppose otherwise.  Write
%%\[I_i = [i_1 <_i i_2 <_i \dots <_i i_r <_i \dots <_i i_k], \quad J = [i_1 <_i i_2 <_i \dots <_i j_r <_i \dots <_i j_k],\]
%%so that $I_i$ and $J$ differ for the first time in the $r$th position.  Then
%%\[r-1 \geq |Prop(\{i_1, \dots, i_{r-1}\})| = |Prop(\{i_1, \dots, i_{r+1},j_r\})|,\]
%%If $|Prop(\{i_1, \dots, i_{r-1}|\}| = r-1$ then we are done; otherwise

@@

However, $j_0$ is the $<_i$-smallest element in the symmetric difference of $J$ and $I_i$, so in particular there is some propagator $p_2$ such that $J(p_2)=I_i(p_1)$.  Now, let $j_2 = I_i(p_2)$ and note that $j_2 \neq J(p_2)$ (since $I_i(p_1)=J(p_2)$), so either $I_i(p_2) \not\in J$ or there is a propagator $p_3$ such that $J(p_3) = I_i(p_2)$.  Continuing likewise we get a list of vertices $j_k$ and propagators $p_k$ such that $J(p_k) = I_i(p_{k-1}) = j_{k-1}$.

We claim that the propagators and vertices in this list are distinct.  The claim is proved by induction.  The definition of $j_0$ gives that $j_0$ and $j_1$ are distinct and that $p_1$ and $p_2$ are distinct.  Now, suppose we have $p_k$ and $j_k$ with $I_i(p_k)=j_k$ and with $j_k\in J$.  Then $p_{k+1}$ is the propagator associated to $j_k$ by the fixed bijection we have chosed for $J$.  Since $j_k$ is distinct from the previous $j_j$, $p_{k+1}$ is distinct from the previous $p_j$.  Then $j_{k+1} = I_i(p_{k+1})$, but $I_i$ is also a bijection, so since $p_{k+1}$ is distinct from the other $p_j$ we have that $j_{k+1}$ is distinct from the other $j_j$ for $1\leq j\leq k$.  Additionally $j_{k+1}\neq j_0$ since $j_0\not\in I_i$ but $j_{k+1}\in I_i$.  This proves the claim.

By finiteness the list must end, and the only way it can end is with some $p_k$ such that $j_k = I_i(p_k)\not\in J$, so in particular $j_k >_i j_0$.

Next we will prove that $j_k <_i j_0$ for all $k > 0$.  This gives a contradiction to the exisitence of $J$ and completes the proof of the theorem.  This proof will also be by induction, however, for the induction to go through nicely we will need to prove the following slightly stronger statement:
for all $k>0$ in the list
\begin{itemize}
\item $j_k <_i j_0$ and
\item either
  \begin{itemize}
  \item $I_i(p_k) <_i J(p_k)$, or
  \item there exists an $ell<k$ such that $I_i(p_\ell) <_i J(p_\ell)$, those two vertices are supported by distinct ends of $p_\ell$ and $p_k$ is on the side of $p_l$ which runs from $I_i(p_\ell)$ to $J(p_\ell)$.
  \end{itemize}
\end{itemize}

We already observed that $j_1 <_i j_0$ and $I_i(p_1) <_k J(p_1)$ so we have the base case for the induction.  Suppose we have the inductive hypothesis for all indices less than $k>1$ and that $p_k$ exists.

By definition $J(p_k) = I_i(p_{k-1}) = j_{k-1}$, so $p_{k-1}$ and $p_{k}$ are both supported at $j_{k-1}$, but $I_{i}(p_{k-1}) = j_{k-1}$.  Thus either $p_{k-1}$ appears before $p_k$ in the neighbourhood of $j_{k-1}$ or $p_k$ was taken earlier by $I_i$ so $I_i(p_k) <_i I_i(p_{k-1})$.  In the latter case we are done as $j_k = I_i(p_k) <_i j_{k-1} <_i j_0$ which is what we want.  So we now assume that $p_{k-1}$ appears before $p_k$ in the neighbourhood of $j_{k-1}$.

>>>>

The entries $c_{p,q}$ are non-zero real variables. The columns of $C(W)$ correspond to the vertices of $W$ listed in their linear order, and the rows correspond to the propagators of $W$. $C(W)$ is therefore only defined up to rearrangement of the rows, which will not matter in this setting since we will only interpret it as parametrizing a family of elements of the $|\cP| \times |V|$ real Grassmannian.

@@

@@

Let $W = (\cP,[n])$ be an admissible Wilson loop diagram with $k$ propagators, and $C(W)$ the associated matrix defined in \eqref{C(W) dfn}. Let $M(W)$ be the matroid realized by $C(W)$, i.e. the matroid with ground set $[n]$ whose independent sets are exactly the sets $V \subseteq [n]$ such that the columns of $C(W)$ indexed by $V$ are linearly independent.

In \cite{wilsonloop}, Agarwala and Amat show that $M(W)$ is also characterized by the Wilson loop diagram itself:

\begin{thm} \label{thm WLD defines matroid} \cite[Theorem 3.6]{wilsonloop} The independent sets of the matroid $M(W)$ associated to an admissible Wilson loop diagram $W = (\cP,[n])$ and realized by $C(W)$ are exactly those subsets $V \subseteq [n]$ such that $\nexists U\subseteq V$ satisfying $|\Prop(U)| < |U|$. \end{thm}
In other words, the independent sets of $M(W)$ correspond to the sets of vertices in $W$ such that no subset supports fewer propagators than the vertices it contains.

By studying $M(W)$ in terms of the behavior of the propagators in $W$, Agarwala and Amat show that $M(W)$ is in fact a positroid whenever $W$ is admissible \cite[Corollary 3.39]{wilsonloop}.

@@

@@

\begin{dfn}\label{props inside p}
Let $W = (\cP, [n])$ be a weakly admissible Wilson loop diagram, with $p \in \cP$ supported on edges $i$ and $j$. Write $(p, <_i)$ to be the same propagator with a direction induced by the linear ordering $<_i$. Then the set of propagators {\bf inside} $(p,<_i)$ is defined to be
\[\cP_{in}(p,<_i) = \{ q \in \cP \ |\ (q,<_i) = (k,l) \text{ and }i \leq_i k <_i l \leq_i j \} \]
and the set of propagators {\bf outside} $(p,<_i)$ is
\[\cP_{out}(p,<_i) = \cP\setminus \cP_{in}(p,<_i).\]
\end{dfn}

@@

Suppose for a contradition that there exists an admissible diagram for which there exists an $i$ with $k\in I_i\setminus \{i\}$ and $k \not\in I_{i+1}$.  Fix $n$.  Let the triple $(W, i, k)$ be such a counterexample on $n$ vertices which is minimal with respect to the number of propagators. %, and among all those counterexamples on $n$ vertices with the same number of propagators, let $(W,i,k)$ be minimal with respect to the distance from $i$ to $k$ in the $\leq _i$ order.

@@

We have shown that $(I_1,\dots,I_n)$ is a Grassmann necklace; it remains to check that this Grassmann necklace corresponds to the positroid $M(W)$.  We need to show that:
\begin{itemize}
\item For each $i \in [n]$, $I_i$ is a basis for $M(W)$.
\item If $J$ is lexicographically smaller than $I_i$ with respect to $<_i$, then $J$ is not a basis for $M(W)$.
\end{itemize}
The algorithm is pairing each $j \in I_i$ with a unique propagator supported on that vertex so by Lemma~\ref{lem basis as perm} it immediately follows that $I_i$ is a basis for $M(W)$.

@@

However, $j_0$ is the $<_i$-smallest element in the symmetric difference of $J$ and $I_i$, so we must have $j_1 \in J$ as well; in particular, there is some propagator $p_2$ such that $J(p_2)=I_i(p_1)$.  Let $j_2 = I_i(p_2)$ and note that $j_2 \neq J(p_2)$ (since $I_i(p_1)=J(p_2)$), so either $I_i(p_2) \not\in J$ or there is a propagator $p_3$ such that $J(p_3) = I_i(p_2)$. Continue this process for as many steps as possible (i.e. for as long as $I_i(p_*) \in J$); this produces a list of vertices $j_k$ and propagators $p_k$ such that 
\[J(p_k) = I_i(p_{k-1}) = j_{k-1}.\]

We claim that the propagators and vertices in this list are distinct; this will show that the process must eventually terminate. The claim is proved by induction.  The definition of $j_0$ gives that $j_0$ and $j_1$ are distinct and that $p_1$ and $p_2$ are distinct.  Now, suppose we have $p_k$ and $j_k$ with $I_i(p_k)=j_k$ and with $j_k\in J$.  Then there must exist a propagator $p_{k+1}$ with $J(p_{k+1}) = j_k$.  Since $J: \cP \longrightarrow [n]$ is injective and $j_k$ is not equal to any of the previous vertices $j_0, \dots, j_{k-1}$ by assumption, $p_{k+1}$ must be distinct from the previous $p_j$.  Define $j_{k+1} = I_i(p_{k+1})$, but since $I_i$ is also a bijection and $p_{k+1}$ is distinct from the earlier $p_j$, it follows that $j_{k+1}$ is distinct from the other $j_j$ for $1\leq j\leq k$.  Additionally $j_{k+1}\neq j_0$ since $j_0\not\in I_i$ but $j_{k+1}\in I_i$.  This proves the claim.

By finiteness the list must end, and the only way it can end is with some $p_k$ such that $j_k = I_i(p_k)\not\in J$. In particular, this implies that $j_k >_i j_0$.

Next we will prove that $j_k <_i j_0$ for all $k > 0$; this gives a contradiction to the existence of $J$ and completes the proof of the theorem.  This proof will also be by induction; however, for the induction to go through nicely we will need to prove the following slightly stronger statement:
for all $k>0$ in the list
\begin{itemize}
\item $j_k <_i j_0$ and
\item either
  \begin{itemize}
  \item $I_i(p_k) <_i J(p_k)$, or
  \item there exists an $\ell<k$ such that $I_i(p_\ell) <_i J(p_\ell)$, those two vertices support distinct ends of $p_\ell$, and $p_k$ is inside $p_l$ when $p_l$ is directed from $I_i(p_\ell)$ to $J(p_\ell)$.
  \end{itemize}
\end{itemize}

We observed above that $j_1 <_i j_0$ and $I_i(p_1) <_i J(p_1)$ so we have the base case for the induction.  Now fix $k > 1$ and suppose that the inductive hypothesis holds for all indices less than $k$ and that $p_k$ exists.

By definition $J(p_k) = I_i(p_{k-1}) = j_{k-1}$, so $p_{k-1}$ and $p_{k}$ are both supported at vertex $j_{k-1}$.

Since $I_i$ always chooses the earliest unassigned propagator supported at a vertex, it follows that either $p_{k-1}$ appears before $p_k$ in the neighbourhood of $j_{k-1}$, or that $p_k$ was assigned earlier than $p_{k-1}$ by the algorithm for $I_i$; note that the ordering $p_1,p_2,\dots,p_k$ need not match the order in which Algorithm~\ref{alg:put GN on WLD} assigns propagators to vertices.

In the latter case we are done, since $p_k$ being assigned earlier than $p_{k-1}$ by $I_i$ means exactly that $I_i(p_k) <_i I_i(p_{k-1})$. Recalling that $I_i(p_{k-1}) = J(p_k) = j_{k-1}$, we obtain $I_i(p_k) <_i J(p_k)$ and $j_k = I_i(p_k) <_i j_{k-1} <_i j_0$, as required.

Now assume that $p_{k-1}$ appears before $p_k$ in the neighbourhood of $j_{k-1}$.

\hlfix{[picture]}{picture}
